---
title: "Quick Start Guide to sapfluxr"
author: "Grant Joyce"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide to sapfluxr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

**sapfluxr** processes and analyses sap flow data from ICT SFM1x heat pulse velocity (HPV) sensors. This guide walks you through the complete workflow from raw data to visualised results.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("neez777/sapfluxr")
```

```{r}
library(sapfluxr)
library(ggplot2)  # For plotting
```

# Basic Workflow

## 1. Import Data

sapfluxr automatically detects and imports multiple data formats from ICT sensors:

```{r eval=FALSE}
# Import raw heat pulse data
heat_pulse_data <- read_heat_pulse_data("path/to/your/data.txt")
```

The import function will:
- Auto-detect the file format (ICT current, legacy, or CSV)
- Parse temperature measurements from all four thermistors (do, di, uo, ui)
- Extract diagnostic information (battery voltage, temperature, etc.)
- Validate data quality (if `validate_data = TRUE`)

```{r include=FALSE}
# Create example data for this vignette
set.seed(42)
n_pulses <- 10

# Simulate realistic temperature data
measurements <- data.frame(
  pulse_id = rep(1:n_pulses, each = 150),
  datetime = rep(seq(as.POSIXct("2024-03-15 08:00:00"),
                     by = "30 min", length.out = n_pulses), each = 150)
)

# Add time within each pulse
measurements$datetime <- measurements$datetime + rep(0:149, times = n_pulses)

# Simulate temperature data with realistic heat pulse response
for (i in 1:n_pulses) {
  pulse_rows <- (i-1)*150 + 1:150
  time_sec <- 0:149

  # Pre-pulse baseline temperatures
  baseline_do <- 18.5 + rnorm(1, 0, 0.1)
  baseline_di <- 18.4 + rnorm(1, 0, 0.1)
  baseline_uo <- 18.6 + rnorm(1, 0, 0.1)
  baseline_ui <- 18.3 + rnorm(1, 0, 0.1)

  # Heat pulse response (rises after 30 sec, peaks around 60-80 sec, decays)
  heat_response <- ifelse(time_sec < 30, 0,
                          exp(-(time_sec - 60)^2 / 800) * runif(1, 0.8, 1.2))

  measurements$do[pulse_rows] <- baseline_do + heat_response * 1.2 + rnorm(150, 0, 0.02)
  measurements$di[pulse_rows] <- baseline_di + heat_response * 1.0 + rnorm(150, 0, 0.02)
  measurements$uo[pulse_rows] <- baseline_uo + heat_response * 0.8 + rnorm(150, 0, 0.01)
  measurements$ui[pulse_rows] <- baseline_ui + heat_response * 0.7 + rnorm(150, 0, 0.01)
}

# Create diagnostic data
diagnostics <- data.frame(
  pulse_id = 1:n_pulses,
  datetime = seq(as.POSIXct("2024-03-15 08:00:00"), by = "30 min", length.out = n_pulses),
  batt_volt = runif(n_pulses, 3.9, 4.1),
  batt_current = c(rnorm(n_pulses-1, 15, 2), rnorm(1, 180, 10)),  # One high heating pulse
  batt_temp = rnorm(n_pulses, 25, 2),
  external_volt = rnorm(n_pulses, 21, 1),
  external_current = c(rnorm(n_pulses-1, 18, 2), rnorm(1, 75, 5))
)

# Create heat_pulse_data object
heat_pulse_data <- list(
  measurements = measurements,
  diagnostics = diagnostics,
  metadata = list(
    file_path = "example_data.txt",
    file_name = "example_data.txt",
    format = "ict_current",
    import_time = Sys.time(),
    file_size = 125000,
    file_size_mb = 0.12,
    n_pulses = n_pulses,
    n_measurements = nrow(measurements),
    chunk_size = 100000,
    r_version = paste(R.version$major, R.version$minor, sep = "."),
    package_version = "0.2.0"
  )
)
class(heat_pulse_data) <- c("heat_pulse_data", "list")
```

### Inspect Imported Data

```{r}
# Summary of imported data
str(heat_pulse_data, max.level = 2)

# View first few temperature measurements
head(heat_pulse_data$measurements)

# View diagnostic information
head(heat_pulse_data$diagnostics)
```

## 2. Calculate Heat Pulse Velocity

Calculate sap velocity using one or more established methods:

```{r}
# Calculate using multiple methods
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "HRMXa", "Tmax_Klu"),
  show_progress = FALSE
)

# View results
head(vh_results)
```

### Available Calculation Methods

| Method | Best For | Reference |
|--------|----------|-----------|
| **HRM** | Low flows, reverse flows | Burgess et al. (2001) |
| **MHR** | Moderate flows | Lopez et al. (2021) |
| **HRMXa/b** | Enhanced HRM accuracy | Burgess & Bleby (unpublished) |
| **Tmax_Coh** | High flows | Cohen et al. (1981) |
| **Tmax_Klu** | High flows (adjusted) | Kluitenberg & Ham (2004) |

**Note:** DMA (Dual Method Approach) is applied as post-processing using `apply_sdma_processing()` - see section below.

### Understanding the Output

```{r}
# Results structure
colnames(vh_results)

# Quality flags
table(vh_results$quality_flag)

# Summary by method
aggregate(Vh_cm_hr ~ method, data = vh_results, FUN = function(x) {
  c(mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    n = sum(!is.na(x)))
})
```

## 3. Visualise Results

### Time Series Plot

```{r fig.width=8, fig.height=5}
# Plot velocity time series
plot_vh_timeseries(
  vh_results,
  methods = c("HRM", "MHR", "HRMXa"),
  sensor_position = "outer"
)
```

### Heat Pulse Trace Plot

Visualise individual heat pulse responses and see exactly where each method calculated its result:

```{r fig.width=8, fig.height=6}
# Plot trace for a single pulse
plot_heat_pulse_trace(
  heat_pulse_data,
  vh_results,
  pulse_id = 5,
  show_methods = c("HRM", "HRMXa", "Tmax_Klu")
)
```

The shaded regions and markers show which time points each method used for its calculation:
- **HRM**: Grey shaded window showing averaging period
- **HRMXa**: Blue shaded window showing averaging period
- **HRMXb**: Green shaded window showing averaging period
- **MHR**: Orange shaded window showing time span from upstream to downstream peak, with orange points and lines marking both upstream and downstream peaks
- **Tmax methods**: Dark green vertical line and point marking time to maximum temperature

This is invaluable for:
- Understanding why methods give different results
- Diagnosing problematic pulses
- Validating calculation windows
- Visualizing the temporal dynamics of heat pulse transport

### Method Comparison

```{r fig.width=8, fig.height=6}
# Compare two methods across all pulses
plot_method_comparison(
  vh_results,
  method1 = "HRM",
  method2 = "MHR",
  sensor_position = "outer"
)
```

## Selectable Dual Method Approach (sDMA)

The **Selectable DMA** (sDMA) is an advanced post-processing method that allows you to choose which high-flow method to pair with HRM. This provides an automatic method selection based on flow conditions.

### How sDMA Works

sDMA automatically switches between two methods based on the **Peclet number (Pe)**:

- **Pe < 1.0**: Uses HRM (optimal for low flows)
- **Pe ≥ 1.0**: Uses your chosen secondary method (optimal for high flows)

The Peclet number is calculated from the HRM velocity: `Pe = (Vh_HRM × probe_spacing) / (diffusivity × 3600)`

### Why Post-Processing?

**New Workflow (Efficient):**
```{r eval=FALSE}
# Step 1: Calculate base methods (each calculated once)
vh <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "Tmax_Klu")
)

# Step 2: Apply sDMA switching as post-processing
vh_sdma <- apply_sdma_processing(vh, secondary_method = "MHR")

# Now have: HRM, MHR, Tmax_Klu, sDMA:MHR (all in one tibble)
```

**Benefits:**
- Each method calculated exactly once (no redundancy)
- Can apply sDMA after corrections: `apply_hpv_corrections() %>% apply_sdma_processing()`
- Can test multiple secondary methods without recalculating HRM
- More explicit about what's happening

### Using sDMA

```{r}
# Step 1: Calculate base methods including HRM (required for Peclet numbers)
vh <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR"),
  show_progress = FALSE
)

# Step 2: Apply sDMA processing with MHR as secondary method
vh_sdma <- apply_sdma_processing(
  vh,
  secondary_method = "MHR",
  show_progress = FALSE
)

# View sDMA-specific columns
head(vh_sdma[vh_sdma$method == "sDMA:MHR",
             c("datetime", "method", "Vh_cm_hr",
               "peclet_number", "selected_method")])
```

The `selected_method` column shows which method was actually used for each measurement (HRM or MHR).

### Compare Multiple sDMA Variants

You can test multiple secondary methods to find which works best for your data:

```{r}
# Calculate base methods once
vh <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "Tmax_Klu"),
  show_progress = FALSE
)

# Apply sDMA with multiple secondary methods at once
vh_multi_sdma <- apply_sdma_processing(
  vh,
  secondary_method = c("MHR", "Tmax_Klu"),
  show_progress = FALSE
)

# Now have: HRM, MHR, Tmax_Klu, sDMA:MHR, sDMA:Tmax_Klu
unique(vh_multi_sdma$method)

# Summary by method
aggregate(Vh_cm_hr ~ method, data = vh_multi_sdma, FUN = mean, na.rm = TRUE)
```

### Visualising sDMA Results

The dedicated sDMA plotting function shows:

- Velocity points colored by which method was selected (HRM vs secondary)
- Peclet number on a secondary y-axis (grey line)
- The Pe = 1.0 switching threshold (dashed horizontal line)
- Legend includes both selected methods and the Peclet number line

```{r fig.width=8, fig.height=6, eval=FALSE}
# Plot sDMA time series with Peclet number
plot_sdma_timeseries(
  vh_sdma,
  sdma_method = "sDMA:MHR",
  sensor_position = "outer"
)

# Plot without Peclet number
plot_sdma_timeseries(
  vh_sdma,
  sdma_method = "sDMA:MHR",
  sensor_position = "outer",
  show_peclet = FALSE
)
```

The plot legend shows which method was used for each data point (and the Peclet number when `show_peclet = TRUE`), and the summary statistics tell you what percentage of measurements used HRM vs the secondary method.

### When to Use sDMA

Use sDMA when:

- **You have variable flow conditions**: Data spans both low and high flow regimes
- **You want method transparency**: sDMA shows exactly which method was used via `selected_method` column
- **Tmax methods fail frequently**: Try `sDMA:MHR` as an alternative to standard DMA (which uses Tmax_Klu)
- **You want to analyse flow regimes**: The `peclet_number` column reveals flow conditions over time

### Advanced: sDMA After Corrections

The post-processing approach allows you to apply sDMA after making probe corrections:

```{r eval=FALSE}
# Calculate base methods
vh <- calc_heat_pulse_velocity(heat_pulse_data, methods = c("HRM", "MHR"))

# Apply corrections (wound diameter, zero-flow, etc.)
vh_corrected <- apply_hpv_corrections(vh, corrections)

# Then apply sDMA to corrected results
vh_sdma <- apply_sdma_processing(vh_corrected, secondary_method = "MHR")
```

# Configurations

## Wood Properties

Wood properties affect thermal calculations. Use built-in configurations or create custom ones:

```{r eval=FALSE}
# Use built-in configurations
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  wood_properties = "eucalyptus"  # or "pine", "generic_sw"
)

# Override specific properties
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  wood_properties = "eucalyptus",
  wood_overrides = list(
    thermal_diffusivity = 0.0028,  # cm²/s
    dbh = 45.2,                    # cm
    sapwood_depth = 3.5            # cm
  )
)
```

### Create Custom Wood Properties

Create a YAML file with your species-specific properties:

```yaml
# wood_my_species.yaml
species:
  common_name: "My Species"
  scientific_name: "Genus species"

thermal_properties:
  thermal_diffusivity:
    value: 0.0025
    units: "cm²/s"
    source: "Measured from cores"

  thermal_conductivity:
    value: 0.0050
    units: "W/(m·K)"

  volumetric_heat_capacity:
    value: 2.0
    units: "MJ/(m³·K)"

physical_properties:
  wood_density:
    value: 550
    units: "kg/m³"

  moisture_content:
    value: 0.40
    units: "fraction"

tree_measurements:
  dbh:
    value: 35.0
    units: "cm"

  sapwood_depth:
    value: 2.8
    units: "cm"
```

Load your custom configuration:

```{r eval=FALSE}
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  wood_properties = "path/to/wood_my_species.yaml"
)
```

## Probe Configuration

Probe configurations define hardware specifications:

```{r eval=FALSE}
# Use default symmetric probe (ICT SFM1x standard)
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  probe_config = "symmetrical"  # Default
)

# Use asymmetric probe (CHPM optimised)
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  probe_config = "asymmetrical"
)
```

### Create Custom Probe Configuration

For custom probe geometries:

```yaml
# probe_custom.yaml
probe_type: "custom_probe"
description: "Custom probe geometry"

geometry:
  probe_spacing:
    value: 0.6
    units: "cm"

  probe_diameter:
    value: 1.5
    units: "mm"

  thermistor_positions:
    downstream_outer: 0.6   # cm from heater
    downstream_inner: 0.3
    upstream_outer: -0.6
    upstream_inner: -0.3

measurement:
  heat_pulse_duration:
    value: 2.0
    units: "seconds"
```

# Advanced Topics

## Custom Calculation Parameters

Override default calculation windows:

```{r eval=FALSE}
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = "HRM",
  parameters = list(
    pre_pulse = 30,      # Baseline period (sec)
    HRM_start = 70,      # HRM window start (sec)
    HRM_end = 110,       # HRM window end (sec)
    L = 0.5,             # HRMX lower bound (proportion of max temp)
    H = 0.8              # HRMX upper bound
  )
)
```

## Filter by Quality

```{r}
# Keep only high-quality results
good_results <- vh_results[vh_results$quality_flag == "OK", ]

# View problematic results
table(vh_results$quality_flag, vh_results$method)
```

## Export Results

```{r eval=FALSE}
# Export to CSV
write.csv(vh_results, "sap_velocity_results.csv", row.names = FALSE)

# Export with metadata
saveRDS(list(
  results = vh_results,
  metadata = heat_pulse_data$metadata,
  processing_date = Sys.time()
), "sap_velocity_complete.rds")
```

# Workflow Summary

```r
# Complete workflow
library(sapfluxr)

# 1. Import
heat_pulse_data <- read_heat_pulse_data("mydata.txt")

# 2. Calculate with multiple methods including sDMA
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "DMA", "sDMA:MHR"),
  wood_properties = "eucalyptus"
)

# 3. Visualise
plot_vh_timeseries(vh_results, methods = c("HRM", "MHR", "DMA"))
plot_heat_pulse_trace(heat_pulse_data, vh_results, pulse_id = 1)

# 3b. Visualise sDMA specifically
plot_sdma_timeseries(vh_results, sdma_method = "sDMA:MHR")

# 4. Export
write.csv(vh_results, "results.csv", row.names = FALSE)
```

## Common Workflows

### Standard Analysis
```r
# Use proven methods for routine analysis
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "DMA")
)
```

### Method Comparison
```r
# Compare multiple methods to find the best for your data
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("HRM", "MHR", "HRMXa", "Tmax_Klu", "DMA")
)
```

### sDMA Optimisation
```r
# If standard DMA has issues, try different secondary methods
vh_results <- calc_heat_pulse_velocity(
  heat_pulse_data,
  methods = c("DMA", "sDMA:MHR", "sDMA:Tmax_Klu", "sDMA:HRMXa")
)

# Visualise each variant
plot_sdma_timeseries(vh_results, sdma_method = "sDMA:MHR")
plot_sdma_timeseries(vh_results, sdma_method = "sDMA:HRMXa")
```

# Getting Help

- **Package documentation**: `?calc_heat_pulse_velocity`
- **Report issues**: https://github.com/neez777/sapfluxr/issues
- **Method details**: See `knowledge_docs/` in package installation

# References

- Burgess, S. S. O., Adams, M. A., Turner, N. C., Beverly, C. R., Ong, C. K., Khan, A. A. H., & Bleby, T. M. (2001). An improved heat pulse method to measure low and reverse rates of sap flow in woody plants. *Tree Physiology*, 21(9), 589-598.

- Cohen, Y., Fuchs, M., & Green, G. C. (1981). Improvement of the heat pulse method for determining sap flow in trees. *Plant, Cell & Environment*, 4(5), 391-397.

- Forster, M. A. (2020). How significant is nocturnal sap flow? *Tree Physiology*, 40(8), 1179-1190.

- Kluitenberg, G. J., & Ham, J. M. (2004). Improved theory for calculating sap flow with the heat pulse method. *Agricultural and Forest Meteorology*, 126(1-2), 169-173.

- Lopez, J. G., Morán-López, T., Benito Garzón, M., & Oliveras Menor, I. (2021). An improved heat pulse method using a single probe for determining sap flow in temperate trees. *Tree Physiology*, 41(12), 2437-2449.
