---
title: "Get Started with sapfluxr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with sapfluxr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette-",
  out.width = "100%"
)
```

# Introduction

**sapfluxr** processes heat pulse velocity (HPV) data from sensors such
as the ICT International SFM1 and SFM1x, taking you from raw logger
temperatures all the way to tree-level water use. Its design philosophy
is built on **transparency**, **non-destructive corrections**, and
**reproducible research**: original measurements are never overwritten,
and every correction adds a new, fully tracked column.

This guide walks the **complete pipeline from start to finish** using
the bundled sample dataset. Each stage is shown with a minimal, runnable
example and a link to a detailed vignette that documents every available
option.

## The processing pipeline

The pipeline mirrors the workflow of the companion `shiny-sapfluxr`
application:

| Stage | What happens | Detailed vignette |
|----------------|------------------------|--------------------------------|
| 1\. Data import | Read and validate raw logger data; optional clock-drift and weather/VPD | [Data Import](vignette-01-data-import.html) |
| 2\. Configuration | Define probe geometry and wood thermal properties | [Probe & Wood Configuration](vignette-02-configuration.html) |
| 3\. HPV calculation | Compute velocities (HRM, MHR, T-max) and flag quality | [HPV Calculation & QC](vignette-03-hpv-calculation.html) |
| 4\. Spacing correction | Identify zero-flow anchors and correct baseline drift | [Baseline & Spacing Correction](vignette-04-spacing-correction.html) |
| 5\. Wound correction | Scale for probe-insertion damage | [Wound Correction](vignette-05-wound-correction.html) |
| 6\. Calibration & sDMA | Align secondary methods and switch on Péclet number | [Calibration & sDMA](vignette-06-calibration-and-sdma.html) |
| 7\. Flux density | Convert to flux density and integrate over sapwood | [Flux Density & Integration](vignette-07-flux-and-integration.html) |
| 8\. Aggregation | Summarise to daily totals and visualise | [Aggregation & Visualisation](vignette-08-aggregation-and-visualisation.html) |

------------------------------------------------------------------------

# A minimal end-to-end pipeline

```{r setup}
library(sapfluxr)
```

## Stage 1: Data import

`read_heat_pulse_data()` auto-detects the sensor format and validates
the data on import.

```{r import}
# Bundled sample dataset (ICT SFM1x format)
hp_file <- system.file("extdata", "Sample_HeatPulse_Data.txt", package = "sapfluxr")
hp_data <- read_heat_pulse_data(hp_file, show_progress = FALSE)

head(hp_data$measurements)
```

→ See [Data Import](vignette-01-data-import.html) for formats,
validation, clock-drift correction, and weather/VPD import.

## Stage 2: Configuration

Heat pulse velocity depends on probe geometry and the thermal properties
of the sapwood. Both are managed through reproducible YAML
configurations.

```{r config}
probe <- load_probe_config("symmetrical")

# Sample wood configuration matching the bundled tree
wood_file <- system.file("extdata", "Sample_Wood_Config.yml", package = "sapfluxr")
wood <- load_wood_properties(wood_file)
```

→ See [Probe & Wood Configuration](vignette-02-configuration.html) for
built-in configs, custom YAML, and deriving thermal properties from core
measurements.

## Stage 3: HPV calculation & quality control

Calculate one or more methods through a single interface, then flag
suspect values.

```{r hpv}
vh_results <- calc_heat_pulse_velocity(
  hp_data,
  methods            = c("HRM", "MHR"),
  probe_config       = probe,
  wood_properties    = wood,
  confirm_parameters = FALSE,
  show_progress      = FALSE
)

vh_results <- flag_vh_quality(vh_results)
head(vh_results)
```

→ See [HPV Calculation & QC](vignette-03-hpv-calculation.html) for the
method physics, baseline options, and quality flags.

## Stage 4: Spacing correction

Identify periods of genuine zero flow (here with the PELT changepoint
detector), then correct baseline drift. `Vs_cm_hr` becomes the **current
best estimate**.

```{r spacing}
anchors <- detect_changepoints(vh_results, sensor_position = "outer")

vh_results <- apply_spacing_correction(
  vh_data         = vh_results,
  changepoints    = anchors$changepoints,
  offset_model    = "segment",
  correction_math = "burgess",
  sensor_position = "both",
  wood_properties = wood,
  verbose         = FALSE
)
```

→ See [Baseline & Spacing
Correction](vignette-04-spacing-correction.html) for PELT, dual-stable,
and VPD anchor detection, plus the offset-model × correction-math
matrix.

## Stage 5: Wound correction

Scale velocities to account for the wound zone created by probe
insertion.

```{r wound}
vh_results <- apply_wound_correction(
  vh_results,
  probe_spacing      = "5mm",
  method             = "linear",
  wood_properties    = wood,
  confirm_parameters = FALSE,
  verbose            = FALSE
)
```

→ See [Wound Correction](vignette-05-wound-correction.html) for linear
vs polynomial models and temporal wound tracking across reinstallations.

## Stage 6: Calibration & sDMA

Align secondary methods to the corrected HRM scale, then switch between
methods based on the Péclet number to capture the full diurnal range.

```{r sdma}
# Calibrate the secondary method(s) to HRM and apply the transform
calibrations <- calibrate_multiple_methods(
  vh_results,
  primary_method    = "HRM",
  secondary_methods = "MHR"
)
vh_results <- transform_multiple_methods(vh_results, calibrations)

# Switch between methods per pulse using the Péclet criterion.
# Pe is computed automatically from probe_config / wood_properties.
vh_results <- apply_sdma_processing(
  vh_results,
  secondary_method = "MHR",
  probe_config     = probe,
  wood_properties  = wood,
  peclet_threshold = 1.0,
  show_progress    = FALSE
)
```

→ See [Calibration & sDMA](vignette-06-calibration-and-sdma.html) for
regression options, switching modes, and the "current best" pointer
column.

## Stage 7: Flux density & integration

Convert velocity to sap flux density ($J_v = Z \cdot V_h$) and integrate
over the sapwood to obtain tree water use ($Q$).

```{r flux}
flux_data <- vh_results
flux_data$Jv_cm3_cm2_hr <- calc_sap_flux_density(
  Vh              = flux_data$Vs_cm_hr,
  wood_properties = wood
)

# Tree dimensions (from the sample wood configuration)
flux_data$dbh                  <- 52.08
flux_data$sapwood_thickness    <- 2.0
flux_data$bark_thickness_dbh   <- 1.2
flux_data$bark_thickness_probe <- 0.5

tree_water_use <- apply_sap_flux_integration(
  flux_data,
  method = "linear_decay"
)
```

→ See [Flux Density &
Integration](vignette-07-flux-and-integration.html) for the Z factor,
sapwood-area calculation, and radial integration models.

## Stage 8: Aggregation & visualisation

Summarise sub-daily flux into daily totals for seasonal water-use
analysis.

```{r daily}
daily_totals <- aggregate_daily(
  tree_water_use,
  total_flux_col = "Q_total_cm3_hr"
)
head(daily_totals)
```

→ See [Aggregation &
Visualisation](vignette-08-aggregation-and-visualisation.html) for
aggregation options, normalised metrics, and the plotting suite.

------------------------------------------------------------------------

# Next steps

This guide is the high-level tour. Each stage links to a detailed
vignette covering all available options, the underlying physics, and
troubleshooting. Start with [Data Import](vignette-01-data-import.html)
and work through the pipeline in order.
