---
title: "Quick Start Guide: The sapfluxr Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide: The sapfluxr Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/vignette-",
  out.width = "100%"
)
```

# Introduction

**sapfluxr** is an open-source R package designed for processing heat pulse velocity (HPV) data from sensors like the ICT International SFM1 and SFM1x. Its core design philosophy is built on **transparency**, **non-destructive corrections**, and **reproducible research**.

This guide walks you through the 10-step processing pipeline implemented in the package, from importing raw sensor data to calculating daily tree-level water use.

## The 10-Step Pipeline

The processing pipeline follows a sequence of physically grounded steps:

1.  **Data Import**: Standardizing raw measurements into a validated structure.
2.  **Wood Property Estimation**: Characterizing the thermal and physical properties of sapwood.
3.  **Heat Pulse Velocity (HPV) Calculation**: Applying multi-method physical equations (e.g., HRM, MHR, T-max).
4.  **Changepoint & Zero-Flow Identification**: Determining temporal calibration periods where flow is minimal.
5.  **Baseline Correction**: Removing zero-flow offsets using segmented, non-destructive models.
6.  **Wound Correction**: Scaling velocity to account for probe insertion damage.
7.  **Secondary Method Calibration**: Aligning methods to ensure a smooth transition across flow regimes.
8.  **Dual Method Switching (sDMA)**: Dynamically selecting the most accurate method for current flow.
9.  **Flux Density & Radial Integration**: Converting velocity to sap flux and scaling to tree-level.
10. **Daily Aggregation**: Integrating sub-daily data into daily totals (L/day).

------------------------------------------------------------------------

# A Minimal Pipeline Example

```{r setup}
library(sapfluxr)
```

## Step 1: Data Import

First, we import the raw data. `sapfluxr` automatically detects the sensor format (CSV, JSON/TXT, or SFT).

```{r import}
# Use a sample file included in the package
sample_file <- system.file("extdata", "sample_ict_current.txt", package = "sapfluxr")
hp_data <- read_heat_pulse_data(sample_file)

# Inspect the structure
head(hp_data$measurements)
```

For more details on supported formats and validation, see [Data Import & Configuration](vignette-01-import-and-config.html).

## Step 2: Wood Property Estimation

Accurate thermal properties are critical. You can load pre-defined configurations or calculate them from core sample measurements.

```{r wood}
# Load a configuration for Eucalyptus
wood_props <- load_wood_properties("eucalyptus")

# View the derived thermal diffusivity (k) used in HPV equations
wood_props$derived_properties$thermal_diffusivity
```

## Step 3: HPV Calculation

`sapfluxr` supports multiple HPV methods. Here we calculate both the **Heat Ratio Method (HRM)** for low flows and **Maximum Heat Ratio (MHR)** for higher flows.

```{r hpv}
# Calculate velocities for multiple methods
vh_results <- calc_heat_pulse_velocity(
  hp_data,
  methods = c("HRM", "MHR"),
  wood_properties = wood_props
)

# View the results (note the Vh_cm_hr_raw column)
head(vh_results)
```

For more details on HPV physics and method selection, see [HPV Calculation Methods](vignette-02-hpv-calculation.html).

## Step 4: Baseline & Spacing Correction

Probes drift over time. We need to identify periods where sap flow is at a genuine zero-flow baseline and correct these offsets. `sapfluxr` separates this into two steps: identifying anchors (e.g., using the **PELT** algorithm to find structural breaks in daily minima), and then applying an offset model and correction math (e.g., **Segmented Burgess spacing correction**).

```{r baseline, eval=FALSE}
# Step 4: Identify zero-flow anchor points
anchors <- detect_changepoints(
  vh_results,
  sensor_position = "outer",
  hpv_method = "HRM",
  penalty = "MBIC"
)

# Step 5: Apply offset model and correction math
vh_results <- apply_spacing_correction(
  vh_data = vh_results,
  changepoints = anchors$changepoints,
  offset_model = "segment",
  correction_math = "burgess",
  sensor_position = "both",
  wood_properties = wood_props
)
```

For more details on baseline identification and correction models, see [Baseline & Zero-Flow Correction](vignette-03-baseline-and-zero-flow.html).

## Step 5: Wound Correction

Wounding at the probe site causes systematic underestimation. We scale the velocity based on the wound diameter.

```{r wound}
# Apply wound correction (B-factor from Burgess et al. 2001)
vh_results <- apply_wound_correction(
  vh_results,
  wound_diameter = 0.2  # cm
)
```

## Step 7 & 8: Calibration & Switching (sDMA)

To capture the full diurnal range, we switch between methods (e.g., HRM and MHR) based on the **Peclet number**. First, we calibrate the secondary method to the corrected HRM scale.

```{r sdma, eval=FALSE}
# 1. Calibrate MHR to HRM scale
vh_results <- calibrate_secondary_method(
  vh_results,
  primary = "HRM",
  secondary = "MHR"
)

# 2. Switch methods (sDMA)
vh_results <- apply_sdma_switching(
  vh_results,
  primary = "HRM",
  secondary = "MHR",
  switching_mode = "peclet",
  threshold = 1.0
)
```

For more details on method integration and sDMA, see [Advanced Corrections & Switching](vignette-04-wound-and-method-correction.html).

## Step 9: Flux Density & Radial Integration

We convert velocity (Vh) to sap flux density (Jv) and integrate over the sapwood area (using the two-annulus model).

```{r flux, eval=FALSE}
# Calculate sapwood areas and convert to tree-level sap flux (Q)
sap_flux <- calc_sap_flux(
  vh_results,
  wood_properties = wood_props,
  sapwood_depth_mm = 35  # example depth
)
```

## Step 10: Daily Aggregation

Finally, we aggregate the sub-daily flux into daily totals.

```{r daily, eval=FALSE}
# Aggregate to daily L/day
daily_totals <- aggregate_daily(
  sap_flux,
  method = "net_sum"
)

head(daily_totals)
```

For more details on scaling and daily aggregation, see [Flux Scaling & Daily Totals](vignette-05-flux-scaling-and-aggregation.html).

------------------------------------------------------------------------

# Next Steps

This quickstart guide provides a high-level overview. For deep dives into each part of the pipeline, explore the detailed vignettes linked above.
