---
title: "8. Aggregation, Metrics & Visualisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{8. Aggregation, Metrics & Visualisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 8** summarises sub-daily flux into daily totals, derives normalised metrics, and
produces the standard diagnostic plots. These are the primary outputs for seasonal water-use
studies and environmental regressions.

# Daily aggregation

`aggregate_daily()` integrates sub-daily measurements (e.g. 15- or 30-minute pulses) into daily
totals, accounting for the measurement interval automatically.

```{r daily, eval=FALSE}
daily_totals <- aggregate_daily(
  tree_water_use,
  datetime_col      = "datetime",
  total_flux_col    = "Q_total_cm3_hr",
  interval          = "auto",
  require_complete_days = FALSE
)
```

## Completeness tracking

The output reports the number of measurements and the data completeness per day, so you can
filter out days with significant gaps (e.g. battery failure) before downstream analysis. Set
`require_complete_days = TRUE` (or `min_measurements_per_day`) to enforce a minimum.

# Normalised metrics

`apply_sapwood_metrics()` derives size-normalised and standardised flux metrics in one step:

* **Mean flux density** — flux per unit sapwood area (`calc_mean_flux_density()`).
* **Normalised flux density** — diurnal pattern standardised to its maximum
  (`normalise_flux_density()`), useful for comparing trees of different sizes.
* **Leaf-area-specific flux** — flux per unit leaf area (`calc_leaf_area_flux()`), if leaf area
  is available.

```{r metrics, eval=FALSE}
metrics <- apply_sapwood_metrics(
  tree_water_use,
  flux_col     = "Q_total_cm3_hr",
  normalise    = TRUE
)
```

`normalise_daily()` provides daily-resolution normalisation for seasonal comparisons.

# Visualisation

All plotting functions use `ggplot2::theme_classic()` as their base. The most useful for this
stage are:

```{r plots, eval=FALSE}
plot_sap_flux_timeseries(tree_water_use)   # sub-daily / hourly series
plot_daily_timeseries(daily_totals)        # daily totals
plot_diurnal_pattern(tree_water_use)       # average diurnal cycle
plot_flux_proportion(flux_data)            # contribution by sensor depth
```

# End of pipeline

You have now followed the full pipeline from raw logger data to daily tree water use. For the
high-level tour, see the [Get Started](sapfluxr.html) guide; to revisit a stage, follow the
links at the foot of each vignette.
