---
title: "7. Flux Density & Sapwood Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. Flux Density & Sapwood Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 7** transforms corrected heat pulse velocity into physically meaningful sap flux
density ($J_v$) and scales point measurements up to whole-tree water use ($Q$).

# Velocity to flux density

Heat pulse velocity measures the speed of heat through the sapwood matrix, which includes both
moving sap and stationary wood. The **sap-flux conversion factor ($Z$)** — derived from wood
density and moisture content in Stage 2 — converts this to sap flux density:

$$J_v = Z \cdot V_h$$

```{r flux_density, eval=FALSE}
flux_data <- vh_results
flux_data$Jv_cm3_cm2_hr <- calc_sap_flux_density(
  Vh              = flux_data$Vs_cm_hr,   # current best velocity
  wood_properties = wood
)
```

# Radial integration

Sap flux varies with depth from the bark, so a sensor's flux density must be integrated over
the annulus of sapwood it represents. `apply_sap_flux_integration()` implements the
two-annulus scheme for SFM1x probes (outer sensor ≈ 7.5 mm, inner sensor ≈ 22.5 mm depth),
weighting each by its sapwood area.

The tree dimensions must be present as columns on the flux data:

```{r dims, eval=FALSE}
flux_data$dbh                  <- 52.08   # cm
flux_data$sapwood_thickness    <- 2.0     # cm
flux_data$bark_thickness_dbh   <- 1.2     # cm (full bark at DBH site)
flux_data$bark_thickness_probe <- 0.5     # cm (remaining bark at probe site)

tree_water_use <- apply_sap_flux_integration(
  flux_data,
  dbh_col                  = "dbh",
  sapwood_thickness_col    = "sapwood_thickness",
  bark_thickness_dbh_col   = "bark_thickness_dbh",
  bark_thickness_probe_col = "bark_thickness_probe",
  method                   = "linear_decay"
)
```

## Integration methods

* **`linear_decay`** — assumes flux density declines linearly between measured depths and toward
  the sapwood–heartwood boundary.
* **`constant_velocity`** — assigns each annulus the flux density of its nearest sensor.

## Sapwood areas and heartwood

`calc_sapwood_areas()` computes the cross-sectional ring areas from DBH, bark thickness, and
sapwood depth. If the sapwood is shallow (e.g. < 22.5 mm), the inner sensor may sit in
non-conducting heartwood; the integration detects this from `sapwood_thickness` and zeroes the
contribution of any sensor beyond the sapwood boundary.

The result carries the per-sensor and combined flux columns `Q_outer_cm3_hr`,
`Q_inner_cm3_hr`, `Q_total_cm3_hr`, `Q_total_L_hr`, and `Q_total_L_day` for whole-tree water use.

# Next step

Continue to [Stage 8: Aggregation & Visualisation](vignette-08-aggregation-and-visualisation.html).
