---
title: "3. HPV Calculation & Quality Control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. HPV Calculation & Quality Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 3** computes heat pulse velocity ($V_h$) from raw temperatures and flags suspect
values. Because no single method captures the full range of sap velocities, `sapfluxr`
implements a suite of complementary algorithms behind one interface.

# The unified interface

`calc_heat_pulse_velocity()` runs one or more methods simultaneously, which makes comparison
and the downstream Selectable Dual Method Approach (sDMA) straightforward.

```{r hpv, eval=FALSE}
vh_results <- calc_heat_pulse_velocity(
  hp_data,
  methods            = c("HRM", "MHR", "Tmax_Klu"),
  probe_config       = probe,
  wood_properties    = wood,
  baseline_method    = NULL,        # use the configured default
  confirm_parameters = FALSE,
  show_progress      = FALSE
)
```

The output is a tibble with `datetime`, `pulse_id`, `method`, `sensor_position`, the raw
velocity `Vh_cm_hr`, and method-specific diagnostic columns. It also carries `prepulse_temp_c`,
the representative pre-pulse baseline temperature per sensor position, which the
Becker & Edwards sap-flux conversion uses downstream (see
[Stage 7](vignette-07-flux-and-integration.html)).

# Implemented methods

## Heat Ratio Method (HRM)
The standard for low and reverse flows. Velocity is computed from the logarithmic ratio of
downstream to upstream temperature rise within a fixed window (typically 60–100 s post-pulse).

* **Bidirectional**: yes (detects reverse flow).
* **Reliable range**: ± 18 cm h⁻¹ (theoretical limit ± 54 cm h⁻¹).
* **Reference**: Burgess et al. (2001).

## Maximum Heat Ratio (MHR)
Uses the peak temperature rise rather than a fixed-window average, extending the measurable
range to higher flows.

* **Bidirectional**: yes.
* **Reliable range**: up to ~130 cm³ cm⁻² h⁻¹ flux density.
* **Reference**: López-Bernal et al. (2021).

## T-max methods (Cohen & Kluitenberg)
Compute velocity from the **time to peak** temperature at the downstream sensor. Accurate at
high flows but cannot detect reverse flow and are unreliable at low flows.

* **`Tmax_Coh`** — Cohen formulation (instantaneous pulse).
* **`Tmax_Klu`** — Kluitenberg formulation (finite pulse duration; corrects a 5–15% high-flow
  underestimate).
* **References**: Cohen et al. (1981); Kluitenberg & Ham (2004).

# Pre-pulse baseline methods

Each velocity calculation subtracts a pre-pulse baseline temperature. The method is set
globally through the analysis configuration:

```{r baseline, eval=FALSE}
# mean_30s (default) | mean_3s | slope_intercept
set_analysis_param("baseline.method", "slope_intercept")
```

* **`mean_30s`** — mean of the full 30 s pre-pulse window (default).
* **`mean_3s`** — mean of the final 3 s before the pulse.
* **`slope_intercept`** — OLS line fitted to the pre-pulse data, evaluated at pulse onset.

# Quality control

`flag_vh_quality()` screens the results for illogical values, outliers, excessive
rate-of-change, and missing pulses.

```{r qc, eval=FALSE}
vh_results <- flag_vh_quality(
  vh_results,
  check_illogical       = TRUE,
  detect_outliers       = TRUE,
  rolling_window        = 11,
  rolling_threshold     = 3,
  detect_rate_of_change = TRUE
)
```

Every record receives a `quality_flag` of `"OK"`, `"WARNING"`, `"ERROR"`, or `"SUSPECT"`,
with detail in `quality_issues`.

# Next step

Continue to [Stage 4: Baseline & Spacing Correction](vignette-04-spacing-correction.html).
