---
title: "03. Baseline & Zero-Flow Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Baseline & Zero-Flow Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(sapfluxr)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Baseline correction is the most critical per-sensor processing decision in the sap flow pipeline. Even minor probe misalignments (e.g., 1-2 mm) can cause 100% error in velocity estimates. `sapfluxr` provides a segmented, non-destructive approach to identify and correct these offsets.

# Step 4: Identification of Zero-Flow Anchor Points

The first task is to determine *when* and *where* the zero-flow baseline has shifted due to growth, probe settling, or physical disturbance. `sapfluxr` offers three approaches:

## 1. Manual & PELT (Statistical)
If weather data is unavailable, you can use the **PELT (Pruned Exact Linear Time)** algorithm to find structural breaks in the daily minimum velocities.

```{r pelt, eval=FALSE}
library(sapfluxr)

# Detect structural shifts in daily minima (outer sensor)
# Input vh_results directly - daily minima are calculated internally
anchors <- detect_changepoints(
  vh_results,
  penalty = "MBIC"
)
```

## 2. Dual-Stable (Physiological & Atmospheric)
When meteorological data is available, the preferred approach is **Dual-Stable detection**. This requires both atmospheric demand (VPD) and the sap flow signal to be low and stable simultaneously, providing the highest standard of evidence for genuine zero-flow.

```{r dual_stable, eval=FALSE}
# Identify periods where BOTH sap flow and VPD are stable
dual_results <- find_dual_stable_periods(
  vh_data        = vh_results,
  weather_data   = vpd_data,
  vh_col         = "Vs_cm_hr", # Use current best estimate
  predawn_window = c(2, 6),
  vpd_threshold  = 0.5,
  vh_threshold   = 2.0
)
```

---

# Step 5: Baseline Correction Application

Once temporal anchors are identified, you must decide *how* to calculate the offset and *which* physical model to apply.

## Offset Model (Shape)
*   **Segment (Step-wise)**: Uses the minimum value within a segment as a constant offset. Best for sudden shifts (e.g., physical disturbance).
*   **Gradient (Continuous Ramp)**: Interpolates a temporal ramp between calibration points. Best for slow, monotonic drift (e.g., radial growth).

## Correction Math (Physics)
*   **Linear (Simple Shift)**: Subtracts the detected offset from all measurements.
*   **Burgess (Physics-based)**: Derives implied probe misalignment from the HRM equation and calculates both a sensitivity correction (slope) and a geometric intercept correction.

```{r segment, eval=FALSE}
# Example 1: Apply Segmented Burgess correction
vh_results <- apply_spacing_correction(
  vh_data = vh_results,
  changepoints = anchors$changepoints,
  offset_model = "segment",
  correction_math = "burgess",
  sensor_position = "both",
  wood_properties = wood_props
)
```

```{r gradient, eval=FALSE}
# Example 2: Apply Gradient Linear interpolation
vh_results <- apply_spacing_correction(
  vh_data = vh_results, 
  changepoints = dual_results$changepoints, 
  offset_model = "gradient",
  correction_math = "linear",
  sensor_position = "both"
)
```

---

# The Non-Destructive Architecture

`sapfluxr` never overwrites your raw data. Each correction creates a new column, allowing you to compare the impact of different strategies:

| Column | Correction Applied |
| :--- | :--- |
| `Vh_cm_hr_raw` | Uncorrected |
| `Vh_cm_hr_sc` | Spacing corrected (diagnostic) |
| `Vs_cm_hr` | **Current Best Estimate** (pointer column) |
| `Vh_cm_hr` | Raw HPV (locked) |

By inspecting these columns, you can audit exactly how your raw measurements were transformed into final estimates.

With a stable baseline established, the next step is **Step 6: Wound Correction**.
