---
title: "03. Baseline & Zero-Flow Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Baseline & Zero-Flow Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Baseline correction is the most critical per-sensor processing decision in the sap flow pipeline. Even minor probe misalignments (e.g., 1–2 mm) can cause 100% error in velocity estimates. `sapfluxr` provides a segmented, non-destructive approach to identify and correct these offsets.

# Step 4: Identification of Zero-Flow Periods

The first task is to determine *when* and *where* the zero-flow baseline has shifted due to growth, probe settling, or physical disturbance. `sapfluxr` offers three approaches:

## 1. Manual & PELT (Statistical)
If weather data is unavailable, you can use the **PELT (Pruned Exact Linear Time)** algorithm to find structural breaks in the daily minimum velocities.

```{r pelt, eval=FALSE}
library(sapfluxr)

# Detect structural shifts in daily minima (outer sensor)
vh_results <- detect_changepoints(
  vh_results,
  method = "pelt",
  penalty = "MBIC"
)
```

## 2. Dual-Stable (Physiological & Atmospheric)
When meteorological data is available, the preferred approach is **Dual-Stable detection**. This requires both atmospheric demand (VPD) and the sap flow signal to be low and stable simultaneously, providing the highest standard of evidence for genuine zero-flow.

```{r dual_stable, eval=FALSE}
# 1. Identify periods where VPD < 0.1 kPa and stable
stable_vpd <- detect_stable_vpd_periods(weather_data, threshold = 0.1)

# 2. Combine with stable sap flow periods
dual_stable <- find_dual_stable_periods(vh_results, stable_vpd)
```

---

# Step 5: Baseline Correction Models

Once temporal segments are identified, you must decide *how* to calculate the offset and *which* physical model to apply.

## Offset Calculation: Flat vs. Gradient
*   **Flat (Step-wise)**: Uses the minimum velocity within a segment as a constant offset. Best for sudden shifts (e.g., physical disturbance).
*   **Gradient (Linear Ramp)**: Interpolates a temporal ramp between calibration points. Best for slow, monotonic drift (e.g., radial growth).

```{r gradient, eval=FALSE}
# Apply a linear temporal ramp between dual-stable points
vh_results <- apply_gradient_offset_correction(vh_results, dual_stable)
```

## Physical Correction Models

### 1. Flat Offset Subtraction
The simplest and most transparent approach: subtract the detected offset from all measurements. Applicable to all HPV methods (HRM, MHR, T-max).

```{r flat, eval=FALSE}
vh_results <- apply_zero_flow_offset(
  vh_results,
  correction_model = "flat"
)
```

### 2. Burgess Spacing Correction (HRM Only)
A physics-based model that derives the implied probe misalignment from the HRM equation and calculates both a **sensitivity correction** (slope) and a **geometric intercept correction**.

```{r burgess, eval=FALSE}
# Apply the Burgess (2001) spacing correction to HRM data
vh_results <- burgess_correction(vh_results)
```

---

# The Non-Destructive Architecture

`sapfluxr` never overwrites your raw data. Each correction creates a new column, allowing you to compare the impact of different strategies:

| Column | Correction Applied |
| :--- | :--- |
| `Vh_cm_hr_raw` | Uncorrected |
| `Vh_cm_hr_zf` | Zero-flow offset (Flat) |
| `Vh_cm_hr_sc` | Spacing corrected (Burgess) |
| `Vh_cm_hr` | **Current Best Estimate** (pointer column) |

By inspecting these columns, you can audit exactly how your raw measurements were transformed into final estimates.

With a stable baseline established, the next step is **Step 6: Wound Correction**.
