---
title: "5. Wound Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Wound Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 5** corrects for the wound zone created when probes are inserted. The damaged,
non-conducting tissue around each probe obstructs heat transport and causes a systematic
**underestimation** of velocity (often 40–60%). The correction scales velocity by a coefficient
$B$ indexed by wound diameter and probe spacing:

$$V_c = B \cdot V_h$$

# Applying the correction

```{r wound, eval=FALSE}
vh_results <- apply_wound_correction(
  vh_results,
  probe_spacing      = "5mm",      # or "6mm"
  method             = "linear",   # or "polynomial"
  wood_properties    = wood,
  confirm_parameters = FALSE
)
```

## Correction models

* **`linear`** — a single linear coefficient. Robust and the most widely used.
* **`polynomial`** — a higher-order ($B = b_0 + b_1 d + b_2 d^2 + \dots$) relationship between
  wound diameter and the correction factor, following Burgess et al. (2001).

The built-in coefficient sets are listed by `list_wound_coefficients()` and depend on
`probe_spacing` (`"5mm"` or `"6mm"`).

## Specifying the wound diameter

If you know the wound diameter directly, pass it with `wound_diameter` (cm); otherwise the
function derives it from the wood configuration's `wound_correction` block (drill-bit diameter
plus healing allowance).

# Temporal wound expansion

Wound diameter is not static — callus growth changes the effective wound size over weeks to
months, and each reinstallation resets it. `calc_wound_diameter()` implements a sawtooth growth
model (linear growth from each installation date) so the correction tracks the true wound size
through time.

```{r wound_diam, eval=FALSE}
calc_wound_diameter(
  drill_bit_diameter_mm = 1.3,
  days_since_drilling   = 120
)
```

Provide installation and reinstallation dates (the sample wood configuration carries
`initial_date` / `final_date` / `final_diameter_mm`) and the correction interpolates the
diameter for every pulse.

# Column tracking

The wound-corrected velocity is written to `Vh_cm_hr_wc`, and `Vs_cm_hr` (the current best
estimate) is updated to point to it.

# Next step

Continue to [Stage 6: Calibration & sDMA](vignette-06-calibration-and-sdma.html).
