---
title: "2. Probe & Wood Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Probe & Wood Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 2** establishes the physical context for every downstream calculation: the geometry of
the probe and the thermal properties of the sapwood. Both are managed through reproducible
YAML configurations so that an analysis is fully documented and repeatable.

# Probe configuration

Heat pulse velocity depends on the spacing and arrangement of the heater and temperature
sensors. Most ICT SFM1x sensors use the symmetrical geometry (probe spacing $x = 0.5$ cm).

```{r probe}
probe <- load_probe_config("symmetrical")
list_available_probe_configs()
```

Built-in options include `symmetrical`, `asymmetrical`, and `symmetrical_6mm`. To build a
custom geometry:

```{r probe_custom, eval=FALSE}
probe <- create_custom_probe_config(
  config_name      = "my_probe",
  sensor_positions = c("outer", "inner"),
  probe_diameter   = 1.3
)
```

You can validate that a probe configuration is compatible with the methods you intend to run:

```{r probe_validate, eval=FALSE}
validate_probe_config(probe, methods = c("HRM", "MHR"))
```

# Wood properties

Velocity-to-flux conversion and thermal corrections depend on wood density, moisture content,
and thermal diffusivity.

## Loading a configuration

The package ships with built-in configurations, and you can load a custom YAML file by path.

```{r wood_builtin}
# Built-in (generic softwood / eucalyptus / pine)
wood_builtin <- load_wood_properties("eucalyptus")

# Custom YAML matching the bundled sample tree
wood_file <- system.file("extdata", "Sample_Wood_Config.yml", package = "sapfluxr")
wood <- load_wood_properties(wood_file)

list_available_wood_properties()
```

## Deriving properties from core measurements

If you have measurements from wood cores, `sapfluxr` derives the thermal diffusivity ($k$) and
the sap-flux conversion factor ($Z$) for you. Two equivalent workflows are supported.

**Workflow A — fresh weight & volume** (the standard gravimetric approach):

```{r wood_calc_a, eval=FALSE}
wood <- calculate_wood_properties(
  fresh_weight = 12.5,   # g
  dry_weight   = 8.2,    # g
  fresh_volume = 15.1    # cm^3
)
```

**Workflow B — dual density** (fresh and dry density; volume cancels out):

```{r wood_calc_b, eval=FALSE}
wood <- calculate_wood_properties(
  fresh_density = 0.82,
  dry_density   = 0.54
)
```

## Tree dimensions

Wood configurations can also carry the tree dimensions needed for sapwood integration
(Stage 7) — `dbh`, `sapwood_thickness`, `bark_thickness_dbh`, and `bark_thickness_probe`. The
sample configuration includes these alongside the thermal properties.

# Next step

With probe and wood configured, continue to
[Stage 3: HPV Calculation & Quality Control](vignette-03-hpv-calculation.html).
