---
title: "1. Data Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Data Import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

This is **Stage 1** of the pipeline (see the [Get Started](sapfluxr.html) guide for the full
tour). The goal is to read raw logger data into a validated structure and, optionally, to
correct clock drift and attach weather data for downstream VPD-based corrections.

# Reading heat pulse data

`read_heat_pulse_data()` is the primary entry point. It auto-detects the format and validates
the data on import.

```{r import}
hp_file <- system.file("extdata", "Sample_HeatPulse_Data.txt", package = "sapfluxr")
hp_data <- read_heat_pulse_data(hp_file, show_progress = FALSE)

names(hp_data)
head(hp_data$measurements)
```

The returned object is a list:

| Element | Contents |
|---------|----------|
| `measurements` | Raw temperature records: `datetime`, `pulse_id`, and the four sensor channels `do`, `di`, `uo`, `ui` |
| `diagnostics` | Battery, current, and sensor diagnostics |
| `metadata` | File information and detected format |
| `validation` | Results of the automated quality checks |

## Supported formats

`read_heat_pulse_data()` detects three formats automatically:

1. **ICT current (JSON/TXT)** — the modern SFM1x logger format, with nested temperature and
   diagnostic arrays.
2. **ICT legacy** — tab-delimited / fixed-width legacy exports.
3. **CSV** — standard CSV exports from ICT's Sap Flow Tool.

You can force a format with the `format` argument if auto-detection is ambiguous.

## Validation and cleaning

On import, `sapfluxr` runs several checks (controlled by `validate_data = TRUE`):

* **Timestamp continuity** — detects and fills gaps in the measurement sequence.
* **Incomplete days** — `trim_incomplete_days = TRUE` removes partial days at the start/end of
  a file (those with fewer than ~23 h of data), which would otherwise bias daily-minima
  baseline calculations.
* **Range checks** — flags illogical temperatures.

```{r cleaning, eval=FALSE}
hp_data <- read_heat_pulse_data(
  hp_file,
  validate_data        = TRUE,
  trim_incomplete_days = TRUE
)
```

## Multiple files

To import and combine several logger files (e.g. one per tree), use
`read_multiple_heat_pulse_data()` with matching `tree_ids`.

```{r multi, eval=FALSE}
hp_multi <- read_multiple_heat_pulse_data(
  file_paths = c("tree1.txt", "tree2.txt"),
  tree_ids   = c("T1", "T2")
)
```

# Clock-drift correction

Logger clocks drift over long deployments. `fix_clock_drift()` linearly rescales the timeline
from a known calibration point (the device time observed versus the true time).

```{r clock, eval=FALSE}
hp_data <- fix_clock_drift(
  hp_data,
  device_time_col       = "datetime",
  observed_device_time  = "2024-12-01 10:05:00",
  observed_actual_time  = "2024-12-01 10:00:00"
)
```

# Weather data and VPD (optional)

Several baseline-correction strategies (Stage 4) use vapour pressure deficit (VPD). Import
weather with `read_weather_data()` and compute VPD with `calc_vpd()`.

```{r weather, eval=FALSE}
wx_file <- system.file("extdata", "Sample_Meteorological_Data.txt", package = "sapfluxr")
weather <- read_weather_data(wx_file, confirm = FALSE)

weather$vpd_kPa <- calc_vpd(
  air_temp_c        = weather$air_temp_c,
  relative_humidity = weather$relative_humidity
)
```

# Next step

With validated data in hand, continue to
[Stage 2: Probe & Wood Configuration](vignette-02-configuration.html).
