---
title: "05. Flux Scaling & Daily Totals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{05. Flux Scaling & Daily Totals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The final steps of the **sapfluxr** pipeline transform corrected heat pulse velocity ($V_h$) into physically meaningful sap flux density ($J_v$) and scale these measurements to whole-tree water use ($Q$).

# Step 9: Flux Density & Radial Integration

## 1. Velocity to Flux Density ($J_v$)
Heat pulse velocity ($V_h$) measures the speed of heat moving through the sapwood, which includes both the moving sap and the stationary wood matrix. To extract the true sap velocity, we apply the **sap flux conversion factor ($Z$)**:

$$J_v = Z \cdot V_h$$

The $Z$ factor is derived from the wood density and moisture content measurements established in Step 2.

## 2. Radial Integration (Hatton 1990)
Because sap flow varies with depth from the bark, a single sensor measurement must be integrated over the representative area (annulus) of the sapwood. `sapfluxr` implements the **Hatton et al. (1990) two-annulus model** for SFM1x probes:

*   **Outer Annulus ($A_{so}$)**: Assigned the flux density from the outer sensor (7.5 mm depth).
*   **Inner Annulus ($A_{si}$)**: Assigned the flux density from the inner sensor (22.5 mm depth).

The boundaries of these annuli are determined by the total sapwood depth of the tree.

```{r flux, eval=FALSE}
library(sapfluxr)

# Calculate tree-level sap flux (Q, in cm³/hr)
sap_flux <- calc_sap_flux(
  vh_results,
  wood_properties = wood_props,
  dbh = 35.2,            # Tree diameter (cm)
  bark_thickness = 1.2,   # Bark thickness (cm)
  sapwood_depth_mm = 45   # Sapwood depth (mm)
)
```

### Handling Heartwood
If the sapwood is shallow (e.g., < 22.5 mm), the inner sensor may sit in non-conducting heartwood. `sapfluxr` automatically detects this based on your `sapwood_depth_mm` input and adjusts the integration zones accordingly, zeroing the contribution of any sensors located beyond the sapwood-heartwood boundary.

---

# Step 10: Daily Aggregation

Finally, sub-daily measurements (e.g., 15-minute or 30-minute pulses) are integrated into daily totals. This is the primary output used for seasonal water-use studies and environmental regressions.

## Integration Method
`sapfluxr` integration accounts for the measurement interval automatically:

```{r daily, eval=FALSE}
# Aggregate to daily L/day
daily_totals <- aggregate_daily(
  sap_flux,
  method = "net_sum",
  min_measurements_per_day = 40  # require at least 40 pulses/day
)
```

### Options:
*   **`net_sum`**: Includes negative velocities. This treats nocturnal reverse flow as a genuine hydraulic redistribution that offsets daytime transpiration.
*   **`zero_floor`**: Zeros all negative values before integration. This assumes negative readings are noise or sensor artefacts rather than real flow.
*   **Completeness Tracking**: The output includes `data_completeness` (0-1) and `n_measurements` columns, allowing you to filter out days with significant gaps (e.g., due to battery failure).

---

# Summary

By following these 10 steps, **sapfluxr** ensures that your raw sensor data is transformed into validated, corrected, and integrated water-use estimates through a transparent and reproducible pipeline.

For more advanced workflows, including batch processing of multiple trees or custom method development, see the [Reference Documentation](index.html).
