---
title: "4. Baseline & Spacing Correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Baseline & Spacing Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 4** is the most critical per-sensor decision in the pipeline. Even a 1–2 mm probe
misalignment can cause large velocity errors, and the apparent spacing drifts over time as the
tree grows and the probes settle. Correction has two parts: **identify** when the zero-flow
baseline shifted, then **apply** a correction between those anchor points.

The architecture is non-destructive — corrections are written to new columns, and `Vs_cm_hr`
becomes the **current best estimate** carried forward.

# Step 1 — Identify zero-flow anchors

`sapfluxr` offers three identification strategies.

## PELT changepoints (statistical)

When weather data is unavailable, use the **PELT** (Pruned Exact Linear Time) algorithm to find
structural breaks in the daily-minimum velocities. Daily minima are computed internally.

```{r pelt, eval=FALSE}
anchors <- detect_changepoints(
  vh_results,
  sensor_position = "outer",
  penalty         = "MBIC"
)
```

You can inspect daily minima directly with `calculate_daily_minima()` and review detections
with `plot_changepoints_interactive()`.

## Dual-stable (physiological + atmospheric)

When weather is available, the preferred approach requires **both** atmospheric demand (VPD)
and the sap-flow signal to be low and stable simultaneously — the strongest evidence of genuine
zero flow.

```{r dual, eval=FALSE}
dual <- find_dual_stable_periods(
  vh_data        = vh_results,
  weather_data   = weather,
  vh_col         = "Vs_cm_hr",
  predawn_window = c(2, 6),
  vpd_threshold  = 0.5,
  vh_threshold   = 2.0
)
```

## VPD changepoints

`detect_vpd_changepoints()` finds shifts in atmospheric demand alone, useful for identifying
suitable correction windows.

# Step 2 — Apply the correction

Two independent choices control the correction.

**Offset model (shape):**

* `segment` — uses the segment minimum as a constant step offset. Best for sudden shifts.
* `gradient` — interpolates a temporal ramp between anchors. Best for slow, monotonic drift.

**Correction maths (physics):**

* `linear` — subtracts the detected offset from all measurements.
* `burgess` — derives the implied probe misalignment from the HRM equation and applies both a
  sensitivity (slope) and geometric (intercept) correction.

```{r apply, eval=FALSE}
# Segmented Burgess correction
vh_results <- apply_spacing_correction(
  vh_data         = vh_results,
  changepoints    = anchors$changepoints,
  offset_model    = "segment",
  correction_math = "burgess",
  sensor_position = "both",
  wood_properties = wood
)

# Alternative: gradient + linear from dual-stable anchors
vh_results <- apply_spacing_correction(
  vh_data         = vh_results,
  changepoints    = dual$changepoints,
  offset_model    = "gradient",
  correction_math = "linear",
  sensor_position = "both"
)
```

# Non-destructive column tracking

| Column | Meaning |
|--------|---------|
| `Vh_cm_hr` | Raw HPV (preserved, never overwritten) |
| `Vh_cm_hr_sc` | Spacing-corrected diagnostic |
| `Vs_cm_hr` | **Current best estimate** carried forward |

# Next step

Continue to [Stage 5: Wound Correction](vignette-05-wound-correction.html).
