---
title: "01. Data Import & Configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Data Import & Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The first two steps of the **sapfluxr** pipeline focus on bringing raw sensor data into R and establishing the physical context (probe and wood properties) required for accurate heat pulse velocity calculations.

# Step 1: Data Import

`sapfluxr` is designed to work seamlessly with ICT International SFM1 and SFM1x sensors. The `read_heat_pulse_data()` function is the primary entry point, providing automatic format detection and validation.

## Supported Formats

The function automatically detects three primary data formats:
1.  **ICT Current (JSON/TXT)**: The modern format for SFM1x loggers, containing nested arrays of temperature and diagnostic data.
2.  **ICT Legacy**: Tab-delimited or fixed-width legacy formats.
3.  **CSV**: Standard CSV exports from ICT's Sap Flow Tool.

```{r import, eval=FALSE}
library(sapfluxr)

# Import and auto-detect format
hp_data <- read_heat_pulse_data("path/to/sfm_data.txt")

# The output is a list containing:
# - measurements: The raw temperature records (do, di, uo, ui)
# - diagnostics: Battery, voltage, and sensor temperature
# - metadata: File info and detected format
# - validation: Results of quality checks
```

## Data Validation & Cleaning

During import, `sapfluxr` performs several automated checks:
*   **Timestamp Continuity**: Detects and fills gaps in the measurement sequence.
*   **Incomplete Days**: Optionally removes partial days at the start/end of a file, which can bias daily minima calculations.
*   **Outlier Detection**: Flags illogical temperature values (e.g., >100°C or < -20°C).

```{r cleaning, eval=FALSE}
# Trim incomplete days (less than 23h of data)
hp_data <- read_heat_pulse_data(
  "path/to/sfm_data.txt",
  trim_incomplete_days = TRUE
)
```

## Clock Drift Correction

Logger clocks can drift over long deployments. `fix_clock_drift()` allows you to linearly stretch or compress the timeline based on observed device time versus actual time.

```{r clock, eval=FALSE}
fixed_data <- fix_clock_drift(
  hp_data,
  observed_device_time = "2024-12-01 10:05:00",
  observed_actual_time = "2024-12-01 10:00:00"
)
```

---

# Step 2: Wood Property Estimation

Heat pulse velocity depends directly on the thermal properties of the sapwood. `sapfluxr` provides a structured way to manage these properties.

## The Configuration System

Configurations are managed via YAML files, ensuring that your analysis is reproducible and your parameters are explicitly documented.

### Loading Pre-defined Properties
The package includes several built-in configurations for common species/wood types:

```{r load_config}
# Load built-in eucalyptus properties
wood_props <- load_wood_properties("eucalyptus")

# View the configuration
# print(wood_props)
```

### Creating Custom Configurations
You can create your own YAML files or specify properties as a list.

```{r custom_config, eval=FALSE}
custom_wood <- create_custom_wood_properties(
  config_name = "My_Tree_01",
  dry_density = 450,
  fresh_density = 520,
  moisture_content = 35
)
```

## Calculating Properties from Measurements

If you have measurements from wood core samples, you can calculate the required thermal parameters (diffusivity $k$, and the $Y$ and $Z$ factors) using two equivalent workflows.

### Workflow A: Fresh Weight & Volume
The standard approach requiring fresh weight, dry weight, and fresh volume.

```{r weight_vol, eval=FALSE}
my_wood <- calculate_wood_properties(
  fresh_weight = 12.5,
  dry_weight = 8.2,
  fresh_volume = 15.1
)
```

### Workflow B: Dual-Density
An optimized approach requiring only fresh and dry density (volume cancels out).

```{r dual_density, eval=FALSE}
my_wood <- calculate_wood_properties(
  fresh_density = 0.82,
  dry_density = 0.54
)
```

## Probe Configuration

Finally, you must specify the physical geometry of your probes (e.g., spacing). Most ICT SFM1x sensors use the symmetrical configuration ($x = 0.5$ cm).

```{r probe}
# Load default symmetrical probe (0.5 cm spacing)
probe_cfg <- load_probe_config("symmetrical")
```

With data imported and physical properties defined, you are ready for **Step 3: HPV Calculation**.
