---
title: "02. HPV Calculation Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. HPV Calculation Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The third step of the **sapfluxr** pipeline is the calculation of heat pulse velocity ($V_h$) from raw temperature measurements. Because no single method accurately captures the full range of sap velocities, `sapfluxr` implements a suite of complementary algorithms.

# The Unified Interface: `calc_heat_pulse_velocity()`

All calculation methods are accessed through a single, consistent interface. This design allows you to calculate multiple methods simultaneously, facilitating comparison and enabling the Dual Method Approach (sDMA) downstream.

```{r example, eval=FALSE}
library(sapfluxr)

# Calculate velocities using HRM, MHR, and T-max Kluitenberg
vh_results <- calc_heat_pulse_velocity(
  hp_data,
  methods = c("HRM", "MHR", "Tmax_Klu"),
  wood_properties = "eucalyptus"
)
```

---

# Implemented Methods

## 1. Heat Ratio Method (HRM)
The standard for low and reverse flows. It calculates velocity from the logarithmic ratio of downstream to upstream temperature increases within a fixed time window (typically 60–100 seconds post-pulse).

*   **Bidirectional**: Yes (detects reverse flow).
*   **Reliable Range**: $\pm$ 18 cm h⁻¹.
*   **Theoretical Limit**: $\pm$ 54 cm h⁻¹.
*   **Key Reference**: Burgess et al. (2001).

## 2. Maximum Heat Ratio (MHR)
An extension of HRM that uses the peak temperature increases rather than a fixed window average. This allows it to capture higher flow rates before the signal dissipates at the upstream sensor.

*   **Bidirectional**: Yes.
*   **Reliable Range**: Up to ~130 cm³ cm⁻² h⁻¹ (sap flux density).
*   **Key Reference**: Lopez-Bernal et al. (2021).

## 3. T-max Methods (Cohen & Kluitenberg)
These methods calculate velocity from the **time-to-peak** temperature at the downstream sensor. They are highly accurate at high flows but cannot detect reverse flow and are unreliable at low flows where peaks are broad and poorly resolved.

*   **Cohen formulation**: Assumes an instantaneous heat pulse.
*   **Kluitenberg formulation**: Accounts for finite pulse duration (e.g., a 2-second pulse), correcting a systematic 5–15% underestimation at high flows.
*   **Bidirectional**: No.
*   **Reliable Range**: ~20 to >100 cm h⁻¹.
*   **Key References**: Cohen et al. (1981); Kluitenberg and Ham (2004).

## 4. Modified HRM (HRMXa/b)
Variants of HRM that use dynamic sampling windows based on a proportion of the peak temperature increase, designed to optimize the signal-to-noise ratio in specific environments.

---

# The Péclet Number (Pe)

For every pulse, `sapfluxr` calculates the dimensionless **Péclet number**:

$$Pe = \frac{V_h \cdot x}{k}$$

Where $x$ is probe spacing and $k$ is thermal diffusivity. The Péclet number quantifies the ratio of advective (convection) to diffusive (conduction) heat transport:
*   **Pe ≤ 1**: Conduction dominates. HRM is the most reliable method.
*   **Pe > 1**: Convection dominates. Secondary methods (MHR, T-max) are preferred.

This physical criterion forms the basis for the **Selectable Dual Method Approach (sDMA)** described in Step 8.

---

# Method Comparison

The `Vh_cm_hr_raw` column in the output allows you to directly compare the performance of different methods on your dataset before any corrections are applied.

```{r comparison, eval=FALSE}
# Compare HRM and MHR outputs
# ggplot(vh_results, aes(x = datetime, y = Vh_cm_hr, color = method)) +
#   geom_line() +
#   facet_wrap(~sensor_position)
```

With raw velocities calculated, the next step is to address probe drift and misalignment in **Step 4: Changepoint & Zero-Flow Identification**.
