---
title: "6. Calibration & sDMA Switching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Calibration & sDMA Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(sapfluxr)
```

# Overview

**Stage 6** combines methods so that the full diurnal range of sap flow is captured
accurately: low nocturnal flows with the precision of HRM, and midday peaks with the extended
range of MHR or T-max. This is a two-part process — **calibrate** the secondary method(s) onto
the corrected HRM scale, then **switch** between methods based on the Péclet number (the
Selectable Dual Method Approach, sDMA).

# Why calibration is needed

When two methods are combined, a "junction artefact" (a step change) often appears at the
switching boundary, because the baseline correction (Stage 4) is derived from HRM and does not
perfectly align with the secondary method's raw scale. Calibration removes this by regressing
the secondary method onto the corrected HRM over their overlapping valid range.

## Comparing methods

`compare_methods_enhanced()` and `compare_methods_segmented()` help you understand the
relationship and find the handover threshold before committing to a calibration.

```{r compare, eval=FALSE}
cmp <- compare_methods_enhanced(
  vh_corrected     = vh_results,
  primary_method   = "HRM",
  secondary_method = "MHR",
  sensor_position  = "outer",
  try_quadratic    = TRUE
)
```

## Calibrating

Calibrate a single method with `calibrate_method_to_primary()`, or all secondary methods at
once with `calibrate_multiple_methods()`, then apply the transforms with
`transform_multiple_methods()`.

```{r calibrate, eval=FALSE}
calibrations <- calibrate_multiple_methods(
  vh_results,
  primary_method    = "HRM",
  secondary_methods = "MHR"
)

vh_results <- transform_multiple_methods(vh_results, calibrations)
```

Calibrated velocities are written to `Vh_cm_hr_<method>_cal` (HRM is left unchanged), with the
fitted `slope`, `intercept`, and $R^2$ recorded as metadata.

# The Péclet number

The **Péclet number** is the dimensionless ratio that decides which method is most reliable
for a given measurement:

$$Pe = \frac{V_h \cdot x}{k}$$

where $V_h$ is heat-pulse velocity (cm s⁻¹), $x$ is probe spacing (cm), and $k$ is thermal
diffusivity (cm² s⁻¹). It quantifies advective versus diffusive heat transport:

* **Pe < 1** — conduction dominates; HRM is most reliable.
* **Pe ≥ 1** — convection dominates; secondary methods (MHR, T-max) capture the full dynamic
  range more accurately.

Péclet numbers are computed at this stage (after all corrections) from the calibrated
velocities (`Vs_cm_hr`) using `recalculate_peclet()`. They are **not** computed at the earlier
HPV stage because they should reflect corrected, post-calibration velocities.

# sDMA switching

With methods on a common scale, `apply_sdma_processing()` computes the Péclet number
automatically and switches between HRM and the secondary method per pulse. Supply the same
`probe_config` and `wood_properties` objects used throughout the pipeline:

```{r sdma, eval=FALSE}
vh_results <- apply_sdma_processing(
  vh_results,
  secondary_method = "MHR",
  probe_config     = probe,
  wood_properties  = wood,
  peclet_threshold = 1.0,
  show_progress    = FALSE
)
```

If you need the Péclet numbers before calling `apply_sdma_processing()` — for instance to
inspect them or plot the switching boundary — run `recalculate_peclet()` first and
`apply_sdma_processing()` will reuse the column:

```{r sdma_manual, eval=FALSE}
vh_results <- recalculate_peclet(
  vh_results,
  probe_config    = probe,
  wood_properties = wood,
  velocity_col    = "Vs_cm_hr"   # uses current best estimate
)

vh_results <- apply_sdma_processing(
  vh_results,
  secondary_method = "MHR",
  peclet_threshold = 1.0,
  show_progress    = FALSE
)
```

## Switching criterion

* **Pe < `peclet_threshold`** (default 1.0) → HRM (low flow; advection negligible).
* **Pe ≥ `peclet_threshold`** → secondary method (high flow; MHR or T-max more accurate).

The switched result is written to `Vh_cm_hr_sdma`, and `method_used` records which method was
selected for each measurement.

# The "current best" pointer

Throughout the corrections, `Vs_cm_hr` always points to the most accurate available estimate.
After sDMA it reflects the switched series, so downstream flux functions automatically use the
best velocity without any manual column bookkeeping.

# Next step

Continue to [Stage 7: Flux Density & Integration](vignette-07-flux-and-integration.html).
